---
title: "running the mdp package"
author: "Melissa Lever"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## About

The Molecular Degree of Perturbation allows you to quantify the heterogeneity of transcriptome data samples. The `mdp` takes data containing at least two classes (control and test) and assigns a score to all samples based on how perturbed they are compared to the controls. Gene perturbation scores are calculated for each gene within each class. 

## Basic usage

Load expression and pheno data:

```{r}
library(mdp)
data(example_data) # expression data has gene names in the rows
data(example_pheno) # pheno data needs a Sample and Class column

head(example_data[,1:5]) 
head(example_pheno) 
```

Run the `mdp`
```{r}
mdp.results <- mdp(data=example_data, pdata=example_pheno, control_lab = "baseline")
```

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
